---
title: "Neural Price equation for inferring eco-evolutionary dynamics"
output: bookdown::html_document2
date: "2023-12-15"
author: "Willem Bonnaffé"
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("f_betteRplots.r")
source("f_bngm.r")
source("f_model_o.r")
source("f_model_p.r")
```

# Methods

## Neural Price framework

We will define the Price equation framework in subsequent versions of this document.

## Case study: rotifers and algae

### System

We consider first a simple ecosystem with an algal prey a rotifer pretador.
Let $A,R \in \mathcal{R}^{+}$ denote the population density of the algae and rotifer respectively.
We also consider the phenotype $Z \in \mathcal{R}^{+}$ of the algae, which is defined as the average size of anti-predator clumps.
For the sake of mathematical convenience, we log-transform and standardise the variables (i.e. $\bar{X} := (\log(X) - \mu_X)/\sigma_X$, where $\mu_X$ and $\sigma_x$ are the mean and standard deviation on the log-scale).

```{r script_load_time_series, echo=FALSE}
##############
## INITIATE ##
##############

## Goal: load data, functions

## Load data
TS = read.table("data/TS_2.csv",sep=";",header=T)

## Extract time steps and columns of interest
selected_time_steps = 1:nrow(TS)
selected_columns = c("t","R","G","B")
TS = TS[selected_time_steps,]
TS = TS[,selected_columns]

## Shorten column names
column_names = c("times","A","Z","R")
colnames(TS) = column_names

## Transform variables
TS[,2] = log(TS[,2])
TS[,3] = log(TS[,3])
TS[,4] = log(TS[,4])

## Normalise time series
TS[,-1] = apply(TS[,-1], 2, function(x) (x-mean(x))/sd(x))

## Data specs
N = ncol(TS) - 1
n = nrow(TS)
```

```{r fig_plot_time_series, echo=FALSE}
######################
## PLOT TIME SERIES ##
######################

## Plot settings
par(mfrow=c(1,1), xaxs="i", yaxs="i")
color_vector = c("","green", "blue", "red")
labels_vector = c("", "Algae Density", "Mean Phenotype", "Rotifer Density")
xlim = c(0,100)
ylim = c(-4.0,3.0)
## Plotting area
plot(TS[,1], TS[,2], cex=0, xlab="Time Step", ylab="State (SU)", bty="n", xaxt="n", yaxt="n", bty="l", xlim=xlim, ylim=ylim)
## Grid
x = seq(xlim[1],xlim[2],10)
y = seq(ylim[1],ylim[2],0.5)
add_axes_and_grid(x, y, alpha=c(10, 0.2), label="")
## Data
for(i in 2:ncol(TS)) lines(TS[,1], TS[,i], type="p", col=adjustcolor(color_vector[i], alpha=0.5), pch=16)
## Legend
legend("bottomright", legend=labels_vector[-1], col=color_vector[-1], lty=1, bty="n")
## End
par(mfrow=c(1,1))

#
###
```

### Model

We define a simple model, using the Price equation to model the dynamics of the phenotype, to represent the eco-evolutionary dynamics of the system:

\begin{equation}
\left \{ \begin{aligned}
\dot{A} &:= f_A(A, Z, R; \Theta_A) \\
\dot{Z} &:= V \frac{\partial f_A}{\partial Z} (A, Z, R; \Theta_A) \\
\dot{R} &:= f_R(A, Z, R; \Theta_R), \\
\end{aligned} \right. (\#eq:model)
\end{equation}

where the notation $\dot{X}$ indicates the temporal derivative of stage variable X.
The ecological dynamics of the algae and rotifer are defined as neural networks $f_A$ and $f_R$, while the dynamics of the phenotype is proportional to the selection gradient $\partial f_A / \partial Z$.
The additive genetic variance $V$ is considered to be constant throughout the time series in this first model. 
The parameter vectors $\Theta_A$ and $\Theta_B$ correspond to the weights and biases of the neural networks.

### Inference

We estimate the parameter vectors by Bayesian neural gradient matching (BNGM).
To do this, we proceed in two steps.
First, we interpolate the time series and compute the expectation of the states and dynamics.
Second, we fit the system to the interpolated states and dynamics.
We introduce regularisation of the objective with Laplace priors on the parameters.

```{r script_demc_observation_model, echo=FALSE}
# ############################
# ## DEMC OBSERVATION MODEL ##
# ############################
# 
# ## Imports
# library(Rcpp)
# 
# ## Import modules
# sourceCpp("f_demc.cpp")
# 
# ## Parameters of observation model
# N = ncol(TS) - 1
# n = nrow(TS)
# K_o = 10
# W_o = 10
# N_o = W_o*3
# sd1_o = 1.0
# sd2_o = 0.25
# rho = 1
# alpha_i = 1
# 
# ## Explanatory and response variable
# t = TS[,1]
# Y = TS[,-1]
# nt = seq(min(t), max(t), (t[2]-t[1])/alpha_i)
# 
# ## Standardise time steps
# t_ = 1:length(t)
# nt_ = seq(min(t_), max(t_), (t_[2]-t_[1])/alpha_i)
# 
# ## Standardise data
# Y_ = Y
# mean_y = apply(Y_, 2, mean)
# sd_y = apply(Y_, 2, sd)
# Y_ = t((t(Y_) - mean_y)/sd_y)
# 
# ## Interpolate time series
# Yhat_o = ddt.Yhat_o = NULL
# Yhat_o_lo = ddt.Yhat_o_lo = NULL
# Yhat_o_hi = ddt.Yhat_o_hi = NULL
# i = 1
# for(i in 1:N)
# {
#     ## Anchor ensembling
#     chain = NULL
#     for(k in 1:K_o)
#     {
#       ## Fit
#       Omega_0 = rnorm(N_o, 0, 0.001)
#       Omega_f = argmax.logPost(t_, Y_[,i], f_o.eval, ddOmega.f_o.eval, Omega_0, sd1_o, sd2_o)
#       # Omega_f = argmax.logMarPost(t_, Y_[,i], f_o.eval, ddOmega.f_o.eval, Omega_0, 1.0)
#       chain = rbind(chain, Omega_f)
#     }
# 
#     # ## Initiate chain
#     # # target = function(x) return(logMarPost(t_, Y_[,i], f_o.eval, x, 1))
#     # target = function(x) return(logPost(t_, Y_[,i], f_o.eval, x, sd1_o, sd2_o))
#     # nIt = 1000
#     # Omega_0 = rnorm(N_o, 0, 0.001)
#     # Omega_0 = argmax.logPost(t_, Y_[,i], f_o.eval, ddOmega.f_o.eval, Omega_0, sd1_o, sd2_o)
#     # 
#     # ## Run chain
#     # chain = DEMCpp(list("dTarget"=target,
#     #                     "Theta_0"=Omega_0,
#     #                     "epsilon"=0.001,
#     #                     "nIt"=nIt))$chainList
#     # chain = chain[round(0.5*nIt):nIt, -1] # Burn
#     # chain = chain[seq(1, nrow(chain), nrow(chain)/100), ] # Thin
# 
#     ## Expectation
#     Yhat_o_ = t(apply(chain, 1, FUN = function(x) f_o.eval(nt_, x)))
#     Yhat_o_mean_ = apply(Yhat_o_, 2, mean)
#     Yhat_o_lo_ = apply(Yhat_o_, 2, quantile, probs=0.1)
#     Yhat_o_hi_ = apply(Yhat_o_, 2, quantile, probs=0.9)
#     ddt.Yhat_o_ = t(apply(chain, 1, FUN = function(x) ddt.f_o.eval(nt_, x)))
#     ddt.Yhat_o_mean_ = apply(ddt.Yhat_o_, 2, mean)
#     ddt.Yhat_o_lo_ = apply(ddt.Yhat_o_, 2, quantile, probs=0.1)
#     ddt.Yhat_o_hi_ = apply(ddt.Yhat_o_, 2, quantile, probs=0.9)
#     # Omega_f = apply(chain, 2, FUN = function(x) f_o.eval(nt_, x))
#     # Omega_f = chain[which.max(chain[,1]),-1]
#     # chainList = list(chain)
#     # Yhat_o_ = f_o.eval(nt_, Omega_f)
#     # ddt.Yhat_o_ = ddt.f_o.eval(nt_, Omega_f)
# 
#     ## Collect
#     Yhat_o = cbind(Yhat_o, Yhat_o_mean_)
#     Yhat_o_lo = cbind(Yhat_o_lo, Yhat_o_lo_)
#     Yhat_o_hi = cbind(Yhat_o_hi, Yhat_o_hi_)
#     ddt.Yhat_o = cbind(ddt.Yhat_o, ddt.Yhat_o_mean_)
#     ddt.Yhat_o_lo = cbind(ddt.Yhat_o_lo, ddt.Yhat_o_lo_)
#     ddt.Yhat_o_hi = cbind(ddt.Yhat_o_hi, ddt.Yhat_o_hi_)
# }
# #
# ###
```

```{r fig_observation_model, echo=FALSE}
#########################
## PLOT INTERPOLATIONS ##
#########################

## Load results observation model
load("obj_results_o.RData")

## Plot settings
par(mfrow=c(2,1), xaxs="i", yaxs="i", oma=c(3,3,0,0), mar=c(2,2,1,1))
color_vector = c("","green", "blue", "red")

## Subplot 1:
## Subplot 1 settings
labels_vector = c("", "Algae", "Phenotype", "Rotifer")
xlim = c(0, 100)
ylim = c(-4.0, 3.0)
## Plotting area
par(xpd=NA)
plot(TS[,1], TS[,2], cex=0, xlab="", ylab="State (SU)", bty="n", xaxt="n", yaxt="n", bty="l", xlim=xlim, ylim=ylim)
par(xpd=F)
## Grid
x = seq(xlim[1], xlim[2], 10)
y = seq(ylim[1], ylim[2], 0.5)
add_axes_and_grid(x, y, alpha=c(10, 0.4), label="")
## Data
for(i in 2:ncol(TS)) polygon(x=c(nt_, rev(nt_)), y=c(Yhat_o_hi[,i-1], rev(Yhat_o_lo[,i-1])), border=NA, col=grey(0.75, alpha=0.5))
for(i in 2:ncol(TS)) lines(TS[,1], TS[,i], type="p", col=adjustcolor(color_vector[i], alpha=0.5), pch=16)
for(i in 2:ncol(TS)) lines(nt_, Yhat_o[,i-1], type="l", col=color_vector[i])
## Legend
legend("bottom", legend=labels_vector[-1], col=color_vector[-1], lty=1, bty="n", horiz=T)
legend("topright", legend="a.", bty="n")

## Subplot 2:
## Sub plot 2 settings
labels_vector = c("", "Algae", "Phenotype", "Rotifer")
xlim = c(0, 100)
ylim = c(-1.0, 1.0)
## Plotting area
par(xpd=NA)
plot(nt_, ddt.Yhat_o[,1], cex=0, xlab="Time Step", ylab="Dynamics", bty="n", xaxt="n", yaxt="n", bty="l", xlim=xlim, ylim=ylim)
par(xpd=F)
lines(xlim, c(0,0), lty=2)
## Grid
x = seq(xlim[1], xlim[2], 10)
y = seq(ylim[1], ylim[2], 0.5)
add_axes_and_grid(x, y, alpha=c(10, 0.2), label="")
## Data
for(i in 2:ncol(TS)) polygon(x=c(nt_, rev(nt_)), y=c(ddt.Yhat_o_hi[,i-1], rev(ddt.Yhat_o_lo[,i-1])), border=NA, col=grey(0.75, alpha=0.5))
for(i in 2:ncol(TS)) lines(nt_, ddt.Yhat_o[,i-1], type="l", col=color_vector[i])
## Legend
legend("bottom", legend=labels_vector[-1], col=color_vector[-1], lty=1, bty="n", horiz=T)
legend("topright", legend="b.", bty="n")

## End
par(mfrow=c(1,1))

#
###
```


```{r script_demc_process_model, echo=FALSE}
# #######################
# ## FIT PROCESS MODEL ##
# #######################
# 
# ## Parameters of process model
# K_p = 3
# W_p = 10
# N_p = 2 * W_p * (2+N)
# sd1_p = 1.0
# sd2_p = c(1, 1, 1)*0.25
# train_split = 0.75
# 
# ## Variables
# X = Yhat_o
# ddt.X = ddt.Yhat_o
# Y = ddt.Yhat_o
# 
# ## Standardise predictive variables
# X_ = X
# mean_x = apply(X_,2,mean)
# sd_x = apply(X_,2,sd)
# X_ = t((t(X_)-mean_x)/sd_x)
# 
# ## Standardise response variable
# Y_ = Y
# mean_y = apply(Y_,2,mean)
# sd_y = apply(Y_,2,sd)
# Y_ = t((t(Y_)-mean_y)/sd_y)
# 
# ## Split train and test
# s_l = 1:round(nrow(X_)*train_split)
# s_t = -s_l
# 
# ## functions
# Yhat = function(X,Omega) f_p.eval(X, Omega)
# ddOmega.Yhat = function(X,Omega) ddOmega.f_p.eval(X, Omega)
# 
# ## fit
# Yhat_p = ddx.Yhat_p = Geber_p = NULL
# Yhat_p_lo = ddx.Yhat_p_lo = Geber_p_lo = NULL
# Yhat_p_hi = ddx.Yhat_p_hi = Geber_p_hi = NULL
# for(i in 1:N)
# {
#     ## Anchor ensembling
#     chain = NULL
#     for(k in 1:K_p)
#     {
#       ## Fit
#       Omega_0 = rnorm(N_p, 0, 0.001)
#       Omega_f = argmax.logPost(X_[s_l,], Y_[s_l, i], Yhat, ddOmega.Yhat, Omega_0, sd1_p, sd2_p[i])
#       chain = rbind(chain, Omega_f)
#     }
# 
#     # ## Initiate chain
#     # target = function(x) return(logPost(X_[s_l,], Y_[s_l, i], f_p.eval, x, sd1_p, sd2_p[i]))
#     # nIt = 10000
#     # Omega_0 = rnorm(N_p, 0, 0.001)
#     # Omega_0 = argmax.logPost(X_[s_l,], Y_[s_l, i], Yhat, ddOmega.Yhat, Omega_0, sd1_p, sd2_p[i])
#     # 
#     # ## Run chain
#     # chain = DEMCpp(list("dTarget"=target,
#     #                     "Theta_0"=Omega_0,
#     #                     "epsilon"=0.001,
#     #                     "nIt"=nIt))$chainList
#     # chain = chain[round(0.5*nIt):nIt, -1] # Burn
#     # chain = chain[seq(1, nrow(chain), nrow(chain)/100), ] # Thin
# 
#     ## Expectation dynamics
#     Yhat_p_ = t(apply(chain, 1, FUN = function(x) f_p.eval(X_, x)))
#     Yhat_p_mean_ = apply(Yhat_p_, 2, mean)
#     Yhat_p_lo_ = apply(Yhat_p_, 2, quantile, probs=0.1)
#     Yhat_p_hi_ = apply(Yhat_p_, 2, quantile, probs=0.9)
# 
#     ## Expectation effects
#     ddx.Yhat_p_ = t(apply(chain, 1, FUN = function(x) t(ddx.f_p.eval(X_, x))))
#     ddx.Yhat_p_mean_ = t(apply(ddx.Yhat_p_, 2, mean))
#     ddx.Yhat_p_lo_ = t(apply(ddx.Yhat_p_, 2, quantile, probs=0.1))
#     ddx.Yhat_p_hi_ = t(apply(ddx.Yhat_p_, 2, quantile, probs=0.9))
#     
#     ## Expectation Geber
#     Geber_p_ = t(apply(chain, 1, FUN = function(x) t(t(ddt.X) * ddx.f_p.eval(X_, x))))
#     Geber_p_mean_ = t(apply(Geber_p_, 2, mean))
#     Geber_p_lo_ = t(apply(Geber_p_, 2, quantile, probs=0.1))
#     Geber_p_hi_ = t(apply(Geber_p_, 2, quantile, probs=0.9))
#     
#     ## Collect
#     Yhat_p = cbind(Yhat_p, Yhat_p_mean_)
#     Yhat_p_lo = cbind(Yhat_p_lo, Yhat_p_lo_)
#     Yhat_p_hi = cbind(Yhat_p_hi, Yhat_p_hi_)
#     #
#     ddx.Yhat_p = cbind(ddx.Yhat_p, t(ddx.Yhat_p_mean_))
#     ddx.Yhat_p_lo = cbind(ddx.Yhat_p_lo, t(ddx.Yhat_p_lo_))
#     ddx.Yhat_p_hi = cbind(ddx.Yhat_p_hi, t(ddx.Yhat_p_hi_))
#     #
#     Geber_p = cbind(Geber_p, t(Geber_p_mean_))
#     Geber_p_lo = cbind(Geber_p_lo, t(Geber_p_lo_))
#     Geber_p_hi = cbind(Geber_p_hi, t(Geber_p_hi_))
# }
# #
# ###
```


```{r fig_process_model, echo=FALSE}
#########################
## PLOT INTERPOLATIONS ##
#########################

## Load results process model 
load("obj_results_p.RData")

## visualise
for(i in 1:N)
{
    ## Plot settings
    par(mfrow=c(3,1), xaxs="i", yaxs="i", oma=c(3,3,0,0), mar=c(2,2,1,1))
    color_vector = c("green", "blue", "red")
    labels_vector = c("Algae", "Phenotype", "Rotifer")
    #
    ## Subplot 1: dynamics
    ## Subplot settings
    xlim = c(0, 100)
    ylim = c(-3.0, 3.0)
    ## Plotting area
    par(xpd=NA)
    plot(nt_, Yhat_p[,i], cex=0, xlab="", ylab="Dynamics (SU)", bty="n", xaxt="n", yaxt="n", bty="l", xlim=xlim, ylim=ylim);
    par(xpd=F)
    lines(xlim, c(0,0), lty=2)
    ## Grid
    x = seq(xlim[1], xlim[2], 10)
    y = seq(ylim[1], ylim[2], 1.0)
    add_axes_and_grid(x, y, alpha=c(10, 0.5), label="")
    ## Data
    polygon(x=c(nt_, rev(nt_)), y=c(Yhat_p_lo[,i], rev(Yhat_p_hi[,i])), border=NA, col=grey(0.75, alpha=0.5))
    lines(nt_, Yhat_p[,i], col=color_vector[i])
    points(nt_, Y_[,i], type="p", col=adjustcolor(color_vector[i], alpha=0.5), pch=16)
    ## Legend
    legend("bottom", legend=labels_vector, col=color_vector, lty=1, bty="n", horiz=T)
    legend("topright", legend="a.", bty="n")
    #
    ## Subplot 2: effects
    ## Subplot settings
    xlim = c(0, 100)
    ylim = c(-4.0, 4.0)
    ## Plotting area
    par(xpd=NA)
    plot(nt_, Yhat_p[,i], cex=0, xlab="", ylab="Effects (SU)", bty="n", xaxt="n", yaxt="n", bty="l", xlim=xlim, ylim=ylim);
    par(xpd=F)
    lines(xlim, c(0,0), lty=2)
    ## Grid
    x = seq(xlim[1], xlim[2], 10)
    y = seq(ylim[1], ylim[2], 0.5)
    add_axes_and_grid(x, y, alpha=c(10, 0.5), label="")
    ## Data
    for(j in 1:N) polygon(x=c(nt_, rev(nt_)), y=c(ddx.Yhat_p_lo[((j-1)*n+1):(j*n),i], rev(ddx.Yhat_p_hi[((j-1)*n+1):(j*n),i])), border=NA, col=grey(0.75, alpha=0.5))
    for(j in 1:N) lines(nt_, ddx.Yhat_p[((j-1)*n+1):(j*n),i], type="l", col=color_vector[j])
    ## Legend
    legend("bottom", legend=labels_vector, col=color_vector, lty=1, bty="n", horiz=T)
    legend("topright", legend="b.", bty="n")
    #
    ## Subplot 3: Contributions
    ## Subplot settings
    xlim = c(0, 100)
    ylim = c(-1.0, 1.0)
    ## Plotting area
    par(xpd=NA)
    plot(nt_, Yhat_p[,i], cex=0, xlab="", ylab="Contributions (SU)", bty="n", xaxt="n", yaxt="n", bty="l", xlim=xlim, ylim=ylim);
    par(xpd=F)
    lines(xlim, c(0,0), lty=2)
    ## Grid
    x = seq(xlim[1], xlim[2], 10)
    y = seq(ylim[1], ylim[2], 0.25)
    add_axes_and_grid(x, y, alpha=c(10, 0.2), label="")
    ## Data
    for(j in 1:N) polygon(x=c(nt_, rev(nt_)), y=c(Geber_p_lo[((j-1)*n + 1):(j*n),i], rev(Geber_p_hi[((j-1)*n + 1):(j*n),i])), border=NA, col=grey(0.75, alpha=0.5))
    for(j in 1:N) lines(nt_, Geber_p[((j-1)*n + 1):(j*n),i], type="l", col=color_vector[j])
    ## Legend
    legend("bottom", legend=labels_vector, col=color_vector, lty=1, bty="n", horiz=T)
    legend("topright", legend="c.", bty="n")

    ## End
    par(mfrow=c(1,1))
}
#
###
```
